---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Azura Liu"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
#install.packages("agricolae")
library(agricolae)
Lake<-read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

class(Lake$sampledate)
Lake$sampledate<-as.Date(Lake$sampledate, format = "%m/%d/%y")
class(Lake$sampledate)

#2
mytheme <- theme_gray(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Depth has no effect on mean lake temperature recorded during July across all lakes.
Ha: Depth does have an effect on mean lake temperature recorded during July across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
library(lubridate)

Lake.July<-
  Lake %>%
  filter(month(sampledate) == 7)%>%
  select(`lakename`, `year4`, `daynum`, `depth`, `temperature_C`)%>%
  drop_na()

#5
Depth.v.Temp<- 
  ggplot(Lake.July, aes(x = temperature_C, y = -depth)) +
  geom_point() +
  xlim(0, 35) +
  scale_x_continuous(position = "top") + 
  scale_y_continuous(position = "right")+
  geom_smooth(method = lm)+
  labs( x = "Temperature", y = "Depth")
print(Depth.v.Temp)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The plot does show a negative correlation where temperature decreases with increasing depth; however the relationship is not exactly linear but (likely) reciprocal.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
m1<-lm(data = Lake.July, temperature_C ~ depth)
summary(m1)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The statistical test shows a p value smaller than 0.05 ( p < 2.2e-16) with a degree of freedom of 9726, meaning that we reject our null hypothesis and accept the alternative hypothesis where depth does have an effect on temperature.The r-squared value suggests that depth explains 73.9% of the variance in lake temperatures.According to the coefficients, our model can be summarized as Temperature = 21.96 - 1.95 * Depth. It means that at surface, the mean lake temperature is 21.96 degree Celcius; as we go deeper, each meter of depth going downwards results in a 1.95 degree decrease in temperature.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
corr<-cor(Lake.July %>% select (daynum:temperature_C))
corrplot::corrplot.mixed(corr, upper = "ellipse")
#depth and temperature clearly correlated.

Temp.AIC <- lm(data = Lake.July, temperature_C ~ year4 + daynum + depth)
step(Temp.AIC)

#Retaining all three variables gives lowest AIC.

#10
Temp.Reg <- lm(data = Lake.July, temperature_C ~ year4 + daynum + depth)
summary(Temp.Reg)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggested to retain all three variables (temperature_C ~ year4 + daynum + depth). The model explains 74.1% of the observed variance.If we only use depth as the explanatory variable as in #7, the variance explained drops down by 0.2% to 73.9%.So the full model does not provide significant improvement.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Lakes.aov<-aov(data = Lake.July, temperature_C~lakename)
summary(Lakes.aov)
  
Lakes.lm<-lm(data = Lake.July, temperature_C ~ lakename)
summary(Lakes.lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, lake name as a variable has a significant P value (F(8, 9719)= 50, p <2e-16), so there is a significant difference in mean temperature among lakes. Therefore we reject our null hypothesis. The linear model method statistics furthur confirms that, because each lake as a variable has a significant p value, meaning they each affects the mean lake temperature.It would be meaningless to interpret the r-squared as we are not trying to explain the temperature solely on lake names.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
Lake.temp.depth<-
  ggplot(Lake.July, aes(x = temperature_C, y = -depth, color = lakename))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = FALSE)+
  xlim(0, 35)+
  scale_x_continuous(position = "top") + 
  scale_y_continuous(position = "right")+
  labs(y = "Depth/m", x = "Temperature/C", color = "Lake Name")

print(Lake.temp.depth)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lakes.aov)
Lake.diff.groups <- HSD.test(Lakes.aov, "lakename", group = TRUE)
Lake.diff.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:Paul Lake and Ward Lake share the same mean temperatures statistically as Peter Lake.There is not any lake that has a mean temperature statistically distinct from all other lakes. Each lake share similar means with at least one other lake.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A t-test would do for comparing just two populations, if we don't use a one-way ANOVA.


